Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-08-16 paper/letter.tex
updated 1sigma values

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ sketches a massive dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of $71^{+17}_{-25}$ km s$^{-1}$ with outflows of
$79^{+167}_{-60}$ km s$^{-1}$) consistent with a dynamical mass on
the order of $5\times 10^9$ solar masses, which is at least $20$ times
larger than the neutral mass hydrogen in the galaxy.   
We argue that a plaussible explanation for the high dynamical mass is
the presence of a super-massive black hole.  
This work demonstrates the importance of considering rotation and
multiphase physics among the possible conditions shaping the
\lya\ spectra of the first galaxies. 
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

The best parameters in the rotation model are a rotational velocity of 
$v_{max}=300^{+xx}_{-xx}$\kms, an optical depth $\log\tau=7^{+xx}_{-xx}$ and a temperature
of $T=1.5^{+xx}_{-xx}\times 10^{4}$K. 
This translates into a column density of $\log N_{HI} /
\mathrm{atoms\ cm}^{-2} =  20.5^{+xx}_{-xx}$.  
This model is also able to constrain the angle between the rotation
axis and the observational line-of-sight to
$\theta=65^{+xx}_{-xx}$$^{\circ}$.

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+17}_{-25}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=79^{+167}_{-60}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  
$P_{cl}=0.72^{+0.20}_{-0.32}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=5.9^{+3.1}_{-3.4}\times 10^{9}$ $M_{\odot}$.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ sketches a massive dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of $71^{+12}_{-12}$ km s$^{-1}$ with outflows of
$55^{+67}_{-37}$ km s$^{-1}$) consistent with a dynamical mass on
the order of two billion solar masses, which is at least $20$ times
larger than the neutral mass hydrogen in the galaxy.   
We argue that a plaussible explanation for the high dynamical mass is
the presence of a super-massive black hole.  
This work demonstrates the importance of considering rotation and
multiphase physics among the possible conditions shaping the
\lya\ spectra of the first galaxies. 
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

The best parameters in the rotation model are a rotational velocity of 
$V_{\rm max}=348^{+75}_{-48}$ \kms, an optical depth $\log_{10}\tau=6.96^{+0.26}_{-0.18}$,  and a temperature of $\log_{10} T/\mathrm {K} = 4.27^{+0.11}_{-0.18}$. 
This translates into a column density of $\log N_{HI} /
\mathrm{atoms\ cm}^{-2} =  20.5^{+xx}_{-xx}$.  
This model is also able to constrain the angle between the plane perpendicular to the rotation axis and the observational line-of-sight to
$\theta = 35.78^{+2.13}_{-1.88}$ degrees.

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+12}_{-12}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=55^{+67}_{-37}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  
$P_{cl}=0.77^{+0.15}_{-0.44}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=2.9^{+2.1}_{-0.9}\times 10^{9}$ $M_{\odot}$.
